\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-09-07 ResponseToReview.tex

They present an interesting result, their conclusions are supported by their data, and the methods they use are appropriate. However, I feel like they need to make some further effort to understand what is happening mechanistically. What is Hopscotch doing in maize vs these sampled teosintes? I feel like the expression analyses and phenotyping experiments they suggest should be part of a later study should be part of this study. At a minimum, qRT-PCR experiments should be performed assessing tb1 expression in their with/without Hopscotch lines. Expression analyses of gt1, te1, and tru1 would be great, since they invoke variation at these loci to explain their results.\\

\res{Extensive expression analysis of individuals with and without the \emph{Hopscotch} insertion at \emph{tb1} and a number of other genes is a good suggestion for follow-up work but would be a substantial undertaking and indeed more appropriate for another paper. }

Another potential avenue, that the authors also suggest, lies in more directed and exhaustive greenhouse experiments. Variation in red and far-red light has been shown to be important in regulating both tb1 (in Sorghum) and gt1 in maize. Growth chamber experiments (with or without supplemental far red light, with or without Hopscotch), coupled to expression analyses in seedlings, seem like they should be possible.  In the greenhouse experiments that were done, phenotyping was incomplete. Apart from tillering, Hopscotch and natural variation at the tb1 locus appears to affect ear architecture and internode length. Was there any evidence for variation in these other traits? In the longer term, it would be nice to see what happens when you introgress these particular parviglumis loci into maize.\\

\res{We did not measure ear architecture or internode length. 
While \emph{tb1} does impact these traits, our primary interest was in the effect of the \emph{Hopscotch} insertion at \emph{tb1} on tillering given that this is an ecologically relevant phenotype that is known to be substantially affected by \emph{Hopscotch} in a maize background.
Both additional greenhouse and expression analyses would indeed be useful avenues of further investigation.
Nonetheless, these are extensive additional experiments which we are unable to include at this time. 
We have, however, added text to the conclusion (Page 15, lines 389-395) of this paper discussing their utility and the need for further work to elucidate the role of \emph{tb1} in branching architecture in teosinte:
``Future studies should examine expression levels of \emph{tb1} in teosinte with and without the \emph{Hopscotch} insertion and further characterize the effects of additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{te1}).  These data, in conjunction with more exhaustive phenotyping, should help to further clarify the ecological significance of the domesticated \emph{tb1} allele in natural populations of teosinte."}

@@

One minor issue: in Fig. 1, it would be nice to see where the Balsas River Basin is in relation to the sampling shown.\\

\res{We have added the Balsas River in a revised version of Figure 1.}

\section*{Response to Reviewer 2} 
5) Although this paper brings up an interesting question regarding the role of tb1 standing variation in teosinte, it is not able to answer it completely.  It presents results that have been published previously about the presence of hopscotch in teosinte, maize, and mexicana.\\

\res{Reviewer 2 is correct in stating that it was known that the Hopscotch was present in some teosinte individuals \citep{Studer2011}; however, previous studies sampled a small fraction of the populations and individuals per population that we present here and thus did not provide comprehensive estimates. 
Our study is not only concerned with the presence or absence of the \emph{Hopscotch} allele in teosinte but also the frequency at which it segregates in a range-wide sample of teosinte.  
This finer level of detail sheds light on the evolutionary history of the \emph{Hopscotch} insertion (\emph{i.e.}, the domestication allele at \emph{tb1}) in teosinte. 
We have added clarifying text to the introduction (Page 5, lines 64-69): ``The effects of the \emph{Hopscotch} insertion have been studied in maize \citep{Studer2011}, and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes \citep{StuderDoebley2012}, but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte. Previous studies have confirmed the presence of the \emph{Hopscotch} in samples of ssp. \emph{parviglumis}, ssp. \emph{mexicana}, and landrace maize; yet little is known about the frequency at which the \emph{Hopscotch} is segregating in natural populations."}

### 2014-09-07 ResponseToReview.tex

One minor issue: in Fig. 1, it would be nice to see where the Balsas River Basin is in relation to the sampling shown.\\

\res{We have added the Balsas River in a revised version of Figure 1.}

\section*{Response to Reviewer 2} 
5) Although this paper brings up an interesting question regarding the role of tb1 standing variation in teosinte, it is not able to answer it completely.  It presents results that have been published previously about the presence of hopscotch in teosinte, maize, and mexicana.\\

\res{Reviewer 2 is correct in stating that it was known that the Hopscotch was present in some teosinte individuals \citep{Studer2011}; however, previous studies sampled a small fraction of the populations and individuals per population that we present here and thus did not provide comprehensive estimates. 
Our study is not only concerned with the presence or absence of the \emph{Hopscotch} allele in teosinte but also the frequency at which it segregates in a range-wide sample of teosinte.  
This finer level of detail sheds light on the evolutionary history of the \emph{Hopscotch} insertion (\emph{i.e.}, the domestication allele at \emph{tb1}) in teosinte. 
We have added clarifying text to the introduction (Page 5, lines 64-69): ``The effects of the \emph{Hopscotch} insertion have been studied in maize \citep{Studer2011}, and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes \citep{StuderDoebley2012}, but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte. Previous studies have confirmed the presence of the \emph{Hopscotch} in samples of \emph{parviglumis}, \emph{mexicana}, and landrace maize; however little is known about the frequency with which the \emph{Hopscotch} is segregating in natural populations."}

@@

8) Page 5, line 12 - 14: You say "The tb1 locus appears to play an important role in the shade avoidance pathway in Zea mays and other grasses and may therefore be crucial to the ecology of teosinte (Kebrom and Brutnell, 2007; Lukens and Doebley, 1999)." Please clarify; do the references belong to the first half of the sentence, supporting tb1 playing a role in Zea mays, or the second half of the sentence, supporting tb1 possibly being crucial to the ecology of teosinte? Which of the two distinct things have actually been studied? If the first, could you tell us how tb1 has been shown to work in maize under shade conditions? If the second, was it actually studied, or merely hypothesized in teosinte?\\

\res{We have moved the references cited here to more clearly reflect the papers' findings.  We have also added text describing the role of \emph{tb1} in the shade avoidance pathway and clarifying the manner in which this could affect fitness in teosinte. Changes can be seen on page 5, lines 74-79: ``The \emph{tb1} locus appears to play an important role in the shade avoidance pathway in \emph{Zea mays} \citep{LukensDoebley1999} and other grasses \citep{KebromBrutnell2007} via changes in expression levels in response to shading. \citet{LukensDoebley1999} introgressed the teosinte \emph{tb1} allele into a maize inbred background and noted that under low density conditions plants were highly tillered, but that under high density, plants showed significantly reduced tillers and grew taller. 
Based on these results we hypothesize that \emph{tb1} may play a role in the ecology of teosinte, especially in populations where it grows at a very high density."}

@@

\res{Our goal in screening for the \emph{Hopscotch} insertion was to assess its taxonomic and geographic distribution to the extent possible given the samples to which we had access.  Reviewer 2 is correct in noting that ssp. \emph{mexicana} is not as prevalent in our paper as ssp. \emph{parviglumis}.  Though our sample of \emph{mexicana} is range-wide, we only had access to seeds from 17 populations.  Frequency of the \emph{Hopscotch} insertion was low in these populations relative to \emph{parviglumis}, an interesting point that we have discussed further on page \mbh{X}, lines \mbh{XX} of the text.  The low frequency of \emph{Hopscotch} in \emph{mexicana} precluded a phenotyping experiment with this subspecies since a very high number of seeds would have to be germinated in order to obtain enough individuals with the \emph{Hopscotch} allele.}

\lev{Matt, I am not seeing where this is in the text. Do you need me to add something to this point?}

10) Materials and Methods:
Page 5, lines 22 - 23: how many individuals per accession? A fixed number, or variable, and depending on what? This can be mentioned so the reader does not have to go to the appendix or supplements.\\

\res{We have added a range of values (1-38 individuals) per population in the methods, pages 6, lines 88-93: ``We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{mexicana} populations, and 86 \emph{parviglumis} populations; ranging from 1-38 individuals per population, with an average of 11 individuals per population for \emph{parviglumis} and \emph{mexicana} and 2 individuals per landrace accession) and assessed the presence or absence of the \emph{Hopscotch} insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article)."}

@@

11) Page 5, Line 28: mention how big the hopscotch element is.  Can it be amplified in one amplicon? Again, this would be useful to know without having to look in appendix or supplements.\\

\res{The sizes of amplicons are now mentioned explicitly in the Materials and Methods (page 6, lines 102-105) following primer sequences: 
``Homozygotes show a single band for absence of the \emph{Hopscotch} element ($\sim$300bp) and two bands for presence of the element ($\sim$5kb, amplification of the entire element, and $\sim$1.1kb, amplification of part of the element), whereas heterozygotes are three-banded (Appendix 2, See Supplemental Materials with the online version of this article)."
The 5kb amplicon including the entire \emph{Hopscotch} could be amplified in a single reaction.}

@@

12) Page 6 line 6: "When only one PCR resolved well, we scored one allele for the individual".  Sorry, what does this mean? That you have one band (one assumes the short one?). If you are talking presence/absence, how do you know the allele?\\

\res{We developed primers for two PCR reactions that would amplify products with size polymorphism characteristic of the two different alleles (\emph{i.e.}, \emph{Hopscotch} vs. no \emph{Hopscotch}). If PCR products were clearly resolved for only one of these reactions, we included one allele for that individual in our data set rather than infer the diploid genotype. We have clarified this in the text on page 6, lines 97-107. : ``Two PCRs were performed for each individual, one with primers flanking the \emph{Hopscotch} (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, {\small 5'-TCGTTGATGCTTTGATGGATGG-3'}; 
HopR, {\small 5'-AACAGTATGATTTCATGGGACCG-3'}; and HopIntR, {\small  5'-CCTCCACCTCTCATGAGATCC-3'} (Appendix 3 and Appendix 4, See Supplemental Materials with the online version of this article). Homozygotes show a single band for absence of the element ($\sim$300bp) and two bands for presence of the element ($\sim$5kb, amplification of the entire element, and $\sim$1.1kb, amplification of part of the element), whereas heterozygotes show all three bands (Appendix 2, See Supplemental Materials with the online version of this article). Since we developed a PCR protocol for each allele, if only one PCR resolved well, we scored one allele for that individual rather than infer the diploid genotype."}

13) Page 7 lines 14 - 15 "These analyses only included populations in which eight or more individuals were sampled."  How many populations did that include?\\

\res{Thirty-two populations were included in these analyses, and we have edited the text to reflect this number on page 6, lines 112-114: ``These analyses only included populations ($n = 32$) in which eight or more individuals were sampled."}

@@

14) Page 7 lines 17 - 26: how many year's worth of environmental data went into your analysis? Can you hypothesize that tb1 differences can be correlated with current climactic data if the changes and subsequent selections must have happened many generations ago?\\

\res{Our climatic data consisted of the 19 bioclim layers commonly used in ecological niche analyses. Full methods for generation of these data can be found at the following url: \url{http://www.worldclim.org/methods}. Briefly, temperature and precipitation were averaged for each global location over a minimum of ten years during the period from 1950-2000.  Soil data were downloaded from the Harmonized World Soil Database (FAOHWSD; \url{http://webarchive.iiasa.ac.at/Research/LUC/External-World-soil-database/HWSD_Documentation.pdf}). This recently published data set drew together global soil maps generated by a world-wide network of scientists over the last several decades and summarized them in GIS data layers that are available as polygon coverages. Data for our sampling sites were extracted from these data sets based on latitude and longitude using the DIVA-GIS platform. Since data for our analyses were either summarized over the last several decades (climatic data) or are likely to be stable over time (soil data), they have potentially influenced current frequencies of the \emph{Hopscotch} insertion in our sampled populations. We have further clarified this on page 7, lines 124-129 of the text : ``Environmental data were obtained from \url{www.worldclim.org} and soil data were downloaded from the Harmonized World Soil Database \citep{FAOHWSD} at \url{www.harvestchoice.org}.  These data represent average values for the last several decades (climatic data) or are likely stable over time (soil data) and therefore represent conditions important for local adaptation of our samples.  Information from these data sets was summarized by principle component analysis following \citet{Pyhajarvi2013}."}

@@

15) Page 7 lines 22 - 23: "We used genotyping and covariance data from Pyhajarvi et al. (2013) for BayEnv" what kind of dataset is this? Teosinte?\\

\res{We have clarified the dataset in the paper on page 7, lines 121-124: ``We used teosinte (ssp. \emph{parviglumis} and ssp. \emph{mexicana}) genotyping and covariance data from \citet{Pyhajarvi2013} for BayEnv, with the \emph{Hopscotch} insertion coded as an additional SNP. SNP data from \citep{Pyhajarvi2013} were obtained using the MaizeSNP50 BeadChip and Infinium HD Assay (Illumina, San Diego, CA, USA)."}

@@

16) Page 8 lines 20 - 21:" We created haplotype blocks using a custom Perl script that grouped SNPs separated by less than 5kb into haplotypes." Does LD data in your parviglumus support 5Kb?\\

\res{While the recombination landscape will vary to some extent across individuals of teosinte and maize, broad-scale patterns are conserved \citep{bauer2013}.  We therefore relied on previous estimates of the decay of linkage disequilibrium based on data from HapMap2 of maize that suggest LD decays over an average distance of 5500bp \citep{Chia2012}; because LD decay is even more rapid in teosinte \citep{Chia2012} we used a conservative haplotype block size of 5kb as an estimate of the extent of linkage disequilibrium in our sample and the physical distance over which SNPs would have to be separated in order to be in weak linkage and not violate assumptions of the STRUCTURE model.}

17) Page 9 lines 2 - 3: ``Plants were watered three times a day by hand and with an automatic drip containing 10-20-10 fertilizer." This sounds odd; by hand or automatically? Or you turned on the drip by hand? In which case, it was not automatic\\

\res{We have edited the text to clarify, page 9, lines 203-205 : ``Plants were watered three times a day with an automatic drip containing 10-20-10 fertilizer, which was supplemented with hand watering on extremely hot and dry days."}

@@

18) Page 9 lines 7 - 8: ``Culm diameter is not believed to be correlated with tillering index or variation at tb1." So, why did you do it? Because of the last sentence in the materials and methods? Perhaps do not mention here that is not correlated; wait until the end of the materials and methods where you seem to explain it.\\

\res{Thank you for the suggestion, we have edited the text accordingly and have added our explanation of the inclusion of culm diameter at the end of the Materials and Methods section, page 10, lines 217-221: ``Additionally, in order to compare any association between \emph{Hopscotch} genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS. Culm diameter is not believed to be correlated with tillering index or variation at \emph{tb1} and is used as our independent trait for phenotyping analyses."}

@@

19) Page 9 lines 9 - 10: "We genotyped individuals for the Hopscotch insertion following the protocols listed above." The PCR protocols or the sequencing protocols?\\

\res{We have clarified in the manuscript that genotypes were obtained using our PCR protocol.}

20) Results:
Page 10 lines 5 - 6: ``Within our parviglumis and mexicana samples we found the Hopscotch insertion segregating in 37 and four populations, respectively"; remind us how many parviglumis and Mexicana samples you had total? So we can see if this comes out to a very different ratio than in maize."\\

\res{We have added ratios to the text, page 10, lines 227-230: ``Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico."}

### 2014-09-07 manuscript.tex
Edits
minor edits

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and phased data from \citet{Pyhajarvi2013}, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map \citep{Gerke2013}. Since STRUCTURE does not account for LD due to physical linkage we created haplotype blocks using a custom Perl script from \citet[][code available at \url{XXX}]{Hufford2013}.  In maize, LD decays over an average distance of 5500bp \citep{Chia2012}; because LD decay is even more rapid in teosinte \citep{Chia2012} we used a conservative haplotype block size of 5kb.  We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

\textbf{Genotyping---}The genotype at the \emph{Hopscotch} insertion was confirmed with two PCRs for 837 individuals of the 1100 screened. Among the 247 maize landrace accessions genotyped, all but eight were homozygous for the presence of the insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico (Fig. \ref{Fig1Map}). Using our \emph{Hopscotch} genotyping, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled eight or more individuals. We found that F\textsubscript{CT} = 0, and levels of F\textsubscript{ST} among populations within each subspecies (0.22) and among all populations (0.23) are similar to those reported genome-wide in previous studies (\citealt{Pyhajarvi2013}; Table \ref{Table1Fst}). Although we found large variation in \emph{Hopscotch} allele frequency among our populations, BayEnv analysis did not indicate a correlation between the \emph{Hopscotch} insertion and environmental variables (all Bayes Factors $<$ 1).

>>>>

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and phased data from \citet{Pyhajarvi2013}, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map \citep{Gerke2013}. Since STRUCTURE does not account for LD due to physical linkage we created haplotype blocks using a custom Perl script from \citet[][code available at \lev{Matt is your code available through your github?} \url{XXX}]{Hufford2013}.  In maize, LD decays over an average distance of 5500bp \citep{Chia2012}; because LD decay is even more rapid in teosinte \citep{Chia2012} we used a conservative haplotype block size of 5kb.  We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

\textbf{Genotyping---}The genotype at the \emph{Hopscotch} insertion was confirmed with two PCRs for 837 individuals of the 1,100 screened. Among the 247 maize landrace accessions genotyped, all but eight were homozygous for the presence of the insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico (Fig. \ref{Fig1Map}). Using our \emph{Hopscotch} genotyping, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled eight or more individuals. We found that F\textsubscript{CT} = 0, and levels of F\textsubscript{ST} among populations within each subspecies (0.22) and among all populations (0.23) are similar to those reported genome-wide in previous studies (\citealt{Pyhajarvi2013}; Table \ref{Table1Fst}). Although we found large variation in \emph{Hopscotch} allele frequency among our populations, BayEnv analysis did not indicate a correlation between the \emph{Hopscotch} insertion and environmental variables (all Bayes Factors $<$ 1).
